---
title: "My R Markdown Primer"
author: "Zach Wunder"
date: "2025-11-20"
output: pdf_document
---

# Introduction

R Markdown mixes narrative, code, and output.  
It makes analysis reproducible.  
You can knit to HTML or PDF and share the result.

# Quick math examples

Inline example: $\alpha + \beta = \gamma$.

Displayed equation:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^n x_i
$$

# Load data and show summary

```{r load-and-summary, echo=TRUE}
# load dataset and show first rows and basic summary
data("mtcars", package = "datasets")
head(mtcars)
summary(mtcars$mpg)
```

# Compute simple statistics

```{r simple-stats, echo=TRUE}
# compute mean and sd of mpg and group means by cyl
mean_mpg <- mean(mtcars$mpg, na.rm = TRUE)
sd_mpg <- sd(mtcars$mpg, na.rm = TRUE)
group_means <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
mean_mpg
sd_mpg
group_means
```

# Make a quick ggplot

```{r plot-mtcars, echo=TRUE, message=FALSE, warning=FALSE}
# create a scatter with trend lines colored by cylinder
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
library(ggplot2)
ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "MPG vs Horsepower by Cylinder",
       x = "Horsepower",
       y = "Miles per Gallon",
       color = "cyl") +
  theme_minimal()
```

# Small base R plot

```{r base-plot, echo=TRUE}
# simple base histogram of mpg
hist(mtcars$mpg,
     breaks = 8,
     main = "Base: Distribution of MPG",
     xlab = "MPG")
```

# Reflection

R Markdown kept code and text together.  
Knitting produced an HTML file with code output and plots.  
I liked that I can edit code and re-knit to update results.  
One tip. Run chunks interactively first to catch errors before knitting.

# Session info

```{r session-info, echo=TRUE}
sessionInfo()
```
